---
title: "Cleaning the Palmer Penguins Raw Data"
author: "Daniel Cusimano"
date: "2023-02-15"
output: html_document
---


# Setup


```{r, include=FALSE, cache=FALSE}
knitr::read_chunk('Processingcode.R')
```

Load needed packages. 


```{r, packages, message = FALSE, warning = FALSE}
```

# Data loading


```{r, loaddata}
```

# Data Under Examination

I am working with the Palmer Penguins dataset. After establishing the correct path and working directory, I used skimr to examine the state of the data. For clarity, I have added the data dictionary at the end of this summary. 

```{r, exploredata}
```


# Cleaning

Taking a moment to examine the skimr summary, a couple of problems become apparent. Under the character variables, I can see that the number of unique species is wrong (listed as 9 whereas the Palmer Penguins dataset is based on 3) and that the Culmen Length (mm), which should be listed as numeric, is present. This is is where I begin the data cleaning process. 

I started with an examination of species. 

###  Species names

Knowing that the Palmer Penguins set assesses only 3 species, I wanted see why my data lists 9. In calling for the unique variables it is apparent that the discrepancy comes from typos in the data set. 

```{r, cleandata1}
```

To solve this problem I used the gsub() and sub() functions. 

```{r, cleandata1.2}
```
With the typos fixed I decided to shorten the species names to their common names. 

```{r, cleandata1.3}
```
### Shortening the Dataset

After a couple rounds of skimr I decided to cut down on some of the unnecessary variables for this analysis. This cleaned things up a bit and saved space. 

For this task I first shortened the variable names followed by subsetting out columns that were not needed. I also decided to change Species, Island, and Sex to Factors here instead of later. This not only really cleaned up the data summary but highlighted the misdesignation of Culmen Length as a character value. 

```{r, dropcolumns}
```

## Continuous data
### Culmen Length

The next apparent issue is the classification of Culmen Length as a character value. I first examined the Culmen Length dataset to see why this might be listed as such. 

```{r, cleandata2}
```

The issue appeared to be the value designated as "missing". This transformed the value (and all others) into character values. The solution was to change "missing" to NA and then coerce all values for Culmen Length to numeric. 



```{r, cleandata2.2}
```


When I skimr again it is clear that this value is now classified appropriately. In the data summary, however, something seemed strange about the distribution of values within Culmen Length. I examined this distribution with a histogram and a bivariate plot with Body Mass. 


```{r, cleandata2.3}
```
There are two noticable issues with the bivariate plot. When examining both plots, it is apparent that there are three major outliers in Culmen Length. Additionally, the bivariate plot shows Body Mass outliers. I tackled the issues with Culmen Length first. In handling the problematic data, I am told that the outliers represent a missed decimal point. Thus, my goal is to remedy the data points instead of removing them. 

```{r, cleandata3}
```
Culmen Length looks much better. Next task was Body Mass. 

### Body Mass

```{r, cleandata4.1}
```

There are penguins with body mass of <100g with all others are over 3000g. With no evidence of this being a typo, the best course of action is to remove the values. Given that Mass is the main size variable I needed to also remove any missing values. This simplified the work a bit as I was able to replace the outlier values with NA and then remove all missing values. 


```{r, cleandata4.2}
```


# Culmen Depth

The next variable I looked at was Culmen Depth. I start here with a histogram.


```{r, Culmendepth}
```

This doesn't look too bad but I'll plot it with Body Mass to be sure. 

```{r, Culmendepth1}
```

This is an interesting distribution. It didn't strike me as an error, but I decided to check why the data looks this way. To do so, my first instinct was to differentiate between the species. For this I used ggplot.

```{r, Culmendepth2}
```

My instincts were on point. The distribution makes much more sense when you can see the groups. I don't believe there is anything that needs editing here. 

Though not surprising, it is also interesting to see that this distribution correlates with species diversity on different islands. 

```{r, Culmendepth3}
```

# Flipper Length

The last morphological variable is Flipper Length. I began again with a histogram and bivariate plot with Body Mass to check the distribution. 

```{r, Flipperlength}
```

This data shows a pretty normal scaling of Flipper Length with Body Mass. There is no need to investigate any further. That said, the histogram does show an interesting dip in the median range. This again can be explained by mapping the species distribution for these values. 

```{r, Flipperlength2}
```

## Isotopic Data

# Nitrogen

The last sets of continuous data in the Palmer Penguins dataset compiles isotopic measures from the three species. 

Beginning with the Nitrogen data, I set a histogram to observe the distribution. Though I didn't expect this metric to scale with Body Mass like the morphological data, I set the bivariate plot to check. This plot, interestingly, revealed a negative correlation with Body Mass.

```{r, Nitrogen}
```


My next thought was that this correlation will more strongly reflect habitat/dietary differences in the populations. Thus, I decided to plot the data with Species instead of Mass.

```{r, Nitrogen2}
```

This data seems to make sense. No editing needed. 

# Carbon

Finally, I ran the same set of tests with the Carbon data as I did the Nitrogen data. 

```{r, Carbon}
```

It was, again, difficult for me to see if this data was behaving in a logical way. To make sure that it was, I again decided to represent it as grouped by species. When plotting it this way, the distribution of data looks normal and therefore required no editing. 

```{r, Carbon2}
```

## Data is Clean

```{r, Distribution}
```

With the last of the variables tested, I believe this data is good to go. 

```{r, Final}
```

